A general and efficient representation of Ancestral Recombination Graphs

### 2022-04-12 paper.tex
Hide ugly hyperref links

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}
\usepackage{hyperref}

@@

Although Hudson's algorithm was accompanied by a graph
schematic~\citep[][Fig. 5]{hudson1983properties}, the term
``Ancestral Recombination Graph'' or ARG, was introduced later by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral, griffiths1997ancestral},
who provided a closely related and complementary treatment.
Where Hudson's algorithm focusses on efficient \emph{simulation}, the original
ARG literature is focused on mathematical results about the underlying stochastic
process. Griffiths and colleagues used the word
``ARG'' to refer to both a specific branching-coalescing stochastic process and
the graph structure derived from this process, but for clarity and following recent usage,
we will use ARG to refer to a graph structure, and refer to the
process as the Griffiths process or the coalescent with recombination
\citep{minichiello2006mapping}.

@@

% NB: should this para go into the previous section? It would be
% clean to start the section with the subsequent para
% starting "The Griffiths ARG encoding is an elegant and mathematically economical description of the CwR"
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
% FIXME: need some proper statistical language here to describe why these
% methods didn't work so well.
These methods met with limited success
due to the overwhelmingly large state space of ARGs, which also has
no simple geometry or neighbourhood structure that an inference or
sampling method could exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
This highlighted the distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
since parsimonious ARGs have no immediate connection to
the stochastic process. Indeed, the ARG instances produced by
the coalescent with recombination are usually highly unparsimonious,
since they contain many invisible recombinations
% Do we need the following caveat?
(although those produced under the SMC may be parsimonious, depending
on the pattern of mutations that is present).

@@

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph upwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we simply follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, which is straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left (and so we cannot emit edges as they are generated).
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough that,
with high probability, the number of concurrent lineages in the ARG remains
much smaller than $N$ at each time. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(3,4)$
and the interval $(x, L)$ with $(3, 5)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $3$ from $5$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
$\{(x, y]\}$ with $(3,5)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes. Each individual (labelled
$I_1$ to $I_8$) therefore has a paternal genome (on the right) and a maternal one
(on the left), although the pedigree is only shown for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{e} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the bare gARG, illustrating
that the patterns of
transmission between genomes can be followed without necessarily knowing
how they are combined within individuals. This ``genome-perspective''
simply depicts the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

Formally, we can define a gARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section). Deriving the local tree at a point $x$
follows largely the same pattern as for an eARG but is somewhat
simpler because coordinate information is directly associated with
edges. Starting with the sample nodes $S$ we traverse
upwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
exists) and move to the next node $p$.

@@

This method for representing recombinations may seem wasteful,
since more than one node is required for each recombination,
and genomic interval information must be stored on every edge.
However, asymptotically the amount of information stored is the
same, and there is much simplicity to be gained by having
a single type of node and by treating all edges in the same way.
More importantly --- as we see in the next section --- we gain the
ability to simplify the ARG by removing nodes without affecting the
local tree topologies, and also the ability to record only the
regions of genome relevant to the final, sample genomes. This last
feature turns out to be the key to efficient processing of the ARG.

[Possibly more stuff here: clarify why we assign coordinates to edges
rather than nodes. Basically lay out all the stuff that we can
talk about \emph{without} going in to the details of
ancestry resolution.]

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}
\usepackage[hidelinks]{hyperref}

@@

Although Hudson's algorithm was accompanied by a graph
schematic~\citep[][Fig. 5]{hudson1983properties}, the term
``Ancestral Recombination Graph'' or ARG, was introduced later by
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
who provided a closely related and complementary treatment.
Where Hudson's algorithm focusses on efficient \emph{simulation}, the original
ARG literature is focused on mathematical results about the underlying stochastic
process. Griffiths and colleagues used the term
``ARG'' to refer to both a particular branching-coalescing stochastic process and
the graph structure produced by this process.
% but for clarity and following recent usage,
% we will use ARG to refer to a graph structure, and refer to the
% process as the Griffiths process or the coalescent with recombination
% \citep{minichiello2006mapping}.

@@

% NB: should this para go into the previous section? It would be
% clean to start the section with the subsequent para
% starting "The Griffiths ARG encoding is an elegant and mathematically economical description of the CwR"
Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
This highlighted the distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping},
since parsimonious ARGs have no immediate connection to
the stochastic process.

@@

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes.
% Each individual (labelled
% $I_1$ to $I_8$) therefore has a maternal and paternal genome,
% although the pedigree is only shown for concreteness, and
% there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{e} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

Formally, we can define a gARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section). Deriving the local tree at a point $x$
follows largely the same pattern as for an eARG but is somewhat
simpler because coordinate information is directly associated with
edges. Starting with the sample nodes $S$ we traverse
rootwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
and move to the next node $p$.

@@

% This method for representing recombinations may seem wasteful,
% since more than one node is required for each recombination,
% and genomic interval information must be stored on every edge.
% However, asymptotically the amount of information stored is the
% same, and there is much simplicity to be gained by having
% a single type of node and by treating all edges in the same way.
% More importantly --- as we see in the next section --- we gain the
% ability to simplify the ARG by removing nodes without affecting the
% local tree topologies, and also the ability to record only the
% regions of genome relevant to the final, sample genomes. This last
% feature turns out to be the key to efficient processing of the ARG.

% [Possibly more stuff here: clarify why we assign coordinates to edges
% rather than nodes. Basically lay out all the stuff that we can
% talk about \emph{without} going in to the details of
% ancestry resolution.]
